Meta-Analysis of Variation in Sport and Exercise Science

### 2023-09-15 paper/Revised Paper - Meta-Analysis of Variation in Sport and Exercise Science.Rmd
Add track changes to revised named paper

---
title: "Meta-Analysis of Variation in Sport and Exercise Science"
subtitle: "Examples of Application Within Resistance Training Research"
author: 
  - James Steele$^a$
  - James Fisher$^a$
  - Dave Smith$^b$
  - Brad Schoenfeld$^c$
  - Yefeng Yang$^d$
  - Shinichi Nakagawa$^d$
date: "14th September 2023"
abstract: "Meta-analysis has become commonplace within sport and exercise science for synthesising and summarising empirical studies. However, most research in the field focuses upon mean effects; particularly the effects of interventions to improve outcomes such as fitness or performance. It is thought that individual responses to interventions vary considerably. Hence, interest has increased in exploring *precision* or *personalised* exercise approaches. Not only is the mean often affected by interventions, but variances may also be impacted. Exploration of variances in studies such as randomised controlled trials (RCTs) can yield insight into interindividual heterogeneity in response to interventions and help determine generalisability of effects. Yet, larger samples sizes than those used for typical mean effects are required when probing variances. Thus, in a field with small samples such as sport and exercise science, exploration of variance through a meta-analytic framework is appealing. Despite the value of embracing and exploring variation alongside mean effects in sport and exercise science it is rarely applied to research synthesis through meta-analysis.We introduce and evaluate different effect size calculations along with models for meta-analysis of variation using relatable examples from resistance training RCTs."
thanks: |
  Address for correspondence: james.steele@solent.ac.uk
  Affiliations:
  $^a$Faculty of Sport, Health, and Social Sciences, Solent University, UK;
  $^b$Research Centre for Musculoskeletal and Sports Medicine, Manchester Metropolitan University, Manchester, UK;
  $^c$Health Sciences Department, CUNY Lehman College, Bronx, New York, USA;
  $^d$Evolution & Ecology Research Centre and School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, New South Wales, Australia
output: 
  bookdown::pdf_document2:
    extra_dependencies: ["flafter"]
bibliography: references.bib
csl: apa.csl
toc: false
header-includes:
   - \usepackage{caption}
   - \captionsetup[figure]{font=small}
---

---

@@

```
# Introduction
Although the quantitative synthesis of results across studies has existed since the 17th century [@plackettStudiesHistoryProbability1958], the modern-day term "meta-analysis" was coined by Gene Glass [-@glassPrimarySecondaryMetaAnalysis1976]. Since that time, the use of meta-analysis as a tool for the synthesis of research in sport and exercise science has increased considerably [@haggerMetaanalysis2022], with resistance training (RT) accounting for a considerable proportion of this growth (figure \@ref(fig:trend-plot)). Accordingly, throughout the paper we use RT studies as a hopefully familiar example for sport and exercise science researchers. However, to begin we provide a conceptual overview of meta-analyses and effect sizes as typically used within sport and exercise science.

@@

There are two popular models[^1] used for meta-analysis: the fixed-effect model and the random-effects model [@borensteinBasicIntroductionFixedeffect2010]. The fixed effect model assumes that there is one true effect size[^2] that each study included in the meta-analysis has estimated and that any differences in the estimates between individual studies are due to only sampling error. This essentially means that there is a single common effect which is fixed across studies and each study takes samples of individuals from the population to estimate this effect. We can express this model in the following formula:

[^1]: To clarify language here for those unfamiliar, the term and concept *model* is used commonly in statistics. A statistical model essentially is specification of what we think the data generating process might be for a given situation. In the context of meta-analyses the data are usually the individual effects that we have extracted from studies i.e., the results of each study. The model, in mathematical formulae, is intended to approximate the processes that we assume led to the generation of the data.

[^2]: *Effect size* is an agnostic term used for a family of statistics which communicate the strength of a given 'effect' resulting from research. This includes descriptive statistics ranging from mean raw values to correlation coeffients and everything in between [@caldwellCaseDefaultEffect2020] including, as we shall see, statistics describing variation.

@@

where $\hat\theta_{i}$ is the $i\textrm{th}$ effect size ($i = 1,2,\cdots,N_{i}$; where $N_{i}$ is the number of studies and thus effect sizes), $\theta$ is the intercept or overall mean (i.e., the fixed-effect), and $m_{i}$ is the sampling error for each effect size normally distributed with $\sigma^2_{m_{i}}$. Studies with smaller standard errors of their effect estimates have smaller sampling errors and so these higher precision estimates are given greater weight in the model. The weighting given to each study is calculated as:

@@

Where $s_{i}$ is the standard deviation for the effect estimate and thus $s^2_{i}$ is the variance. It is referred to as inverse-variance weighting. Then, the overall weighted mean effect estimate from the model is then calculated as:

@@

Contrastingly, the random effects model does not make the assumption that there is only one fixed-effect. Instead it allows for the true effects that each study estimates to differ. Each study may share a common underlying effect size but, due to differences between studies in factors such as population characteristics, the manner in which outcomes are operationalised, or subtle differences in intervention and context to name a few, it is possible that the actual effect being estimated by each study differs. The assumption of the random-effects model then is that the studies included estimate effects that come from a larger population of effects determined by the inclusion criteria for studies included. So, the model assumes that the studies included reflect a random sample of all possible permutations of study and the effects they estimate from this population distribution of studies and effects. Hence, in the fixed-effect model there is one effect and it is assumed to be *fixed* across studies, whereas in the random-effects model there are many and we examine an assumed *random* sample of them. We can express this model in the following formula:

@@

where $\hat\theta_{i}$ is the $i\textrm{th}$ effect size ($i = 1,2,\cdots,N_{i}$; where $N_{i}$ is the number of studies and thus effect sizes), $\theta_i$ is the intercept or overall mean of the effects across each study, $\tau_{i}$ is the deviation from $\theta_{i}$ for the $i\textrm{th}$ study (i.e., the random-effects), and $m_{i}$ is the sampling error for each effect size normally distributed with $\sigma^2_{m_{i}}$. Essentially this model assumes that each individual study estimates an effect and there will be some sampling error in estimating it. But, the effects each study estimates comes from an overall distribution of true effects with a mean value (i.e., $\theta_{i}$). The weighting in any model employing random effects such as this requires a different approach as the variance of the distribution of the effect sizes the model assumes, known as $\tau^2$, must also be estimated[^3]. This essentially describes the heterogeneity between the effects included. Once this has been estimated we can calculate a weight that is adjusted for the random effects for each effect as:

[^3]: This estimation can be done using a variety of methods and is an area of ongoing investigation as to how different methods perform. This is beyond the scope of this paper to discuss. We note however that the models we present all utilise Restricted Maximum Likelihood estimation.

@@

Then the average of the distribution of effects can be calculated as per equation \@ref(eq:fixed-average-eq) substituting $w_{i}$ for $w^*_{i}$.

Although historically fixed effect models were commonplace in the field of sport and exercise science [@haggerMetaanalysisSportExercise2006] nowadays the random effects model is more often employed [@haggerMetaanalysis2022]. This is likely due to the fact that direct (i.e., exact) replication of studies is rare[^4] and instead studies often explore similar effects across varying moderating factors such as those noted above.

[^4]: Hence current efforts to conduct direct replications (see https://ssreplicationcentre.com/).

As with many other fields [@nakagawaMetaanalysisVariationEcological2015; @usuiMetaanalysisVariationSuggests2021; @millsDetectingHeterogeneityIntervention2021] likely the most common aim in meta-analysis in sport and exercise science, and indeed primary empirical research too, is to estimate the *effect* of an independent variable upon some dependent outcome variable. The dependent variable is often the mean of a measurement and the independent variable is often a categorical grouping; for example, the comparison of an intervention group(s) and a control group, the comparison of intervention groups between one another, or comparison between non-manipulated categories such as biological sex. Indeed, a recent umbrella review [@bernardez-vazquezResistanceTrainingVariables2022] of meta-analyses in RT identified 14 studies examining the manipulation of RT intervention variables (i.e., the comparison of one intervention to another whereby a variable in the intervention was manipulated) on hypertrophy outcomes, all of which focused on the comparison of mean changes between different intervention groups.

Often, due to the varying operationalisations used for broad outcome concepts[^5], an effect size is used which is "standardised" across studies to allow for their synthesis. Most commonly, a magnitude based[^6] effect size statistic [@caldwellCaseDefaultEffect2020], the standardised mean difference (SMD), is used to compare means between groups or conditions. This statistic is usually a version of Cohen's *d* [@cohenStatisticalPowerAnalysis1988], or its bias-corrected[^7] metric referred to as Hedges' *g* [@hedgesStatisticalMethodsMetaAnalysis2014; @borensteinIntroductionMetaAnalysis2021; @nakagawaEffectSizeConfidence2007][^8]. The SMD, and its sampling variance, $s^2_{SMD}$ are given by:

[^5]: For example, *strength* might be examined in different studies using different operationalisations including one repetition maximum testing or maximum voluntary contractions. Or the same operationalisations may be employed but different exercises such as the squat or bench press.

@@

[^7]: For those unfamiliar with the terminology, an estimator for a statistic is unbiased if it produces parameter estimates that are on average correct. Thus a bias corrected statistic is one which would be biased without the correction applied, but otherwise has been shown to be unbiased.

[^8]: We will refer to both merely as the SMD throughout the manuscript for simplicity and note that throughout when reporting a 'SMD' we are reporting the bias-corrected version. We also note that another magnitude based effect size, Glass' $\Delta$, is commonly recommended as it is the simplest form of SMD though makes assumptions about the impact of the intervention having no effect on the denominator (i.e., variance; @caldwellCaseDefaultEffect2020).

@@

Comparison of heterogeneity in responses, such as post-scores or change scores to interventions, are not the only possible use of statistical methods for comparing variances. For example, in other fields such as ecology there have been calls to shift focus of analysis onto the exploration of dispersion of traits between groups in non-experimental or intervention designs [@nakagawaMeanStrikesBack2012]. Some recent examples from sport and exercise science, and RT in particular, include primary research exploring between-participant acute response variation for the purposes of identifying methods[^9] to reduce RT stimulus heterogeneity, [@exnerDoesPerformingResistance2022] as well as a meta-analysis exploring between-participant heterogeneity of accuracy in predicting proximity to task failure during RT [@halperinAccuracyPredictingRepetitions2022] and in the number of repetitions that can be performed at different percentages of one repetition maximum [@nuzzoMaximalNumberRepetitions2023].

@@

Although ultimately we will recommend the thoughtful consideration of assumptions for certain approaches regarding what effect size statistics and models to employ in examining variation through use of meta-analysis, we provide examples throughout using all approaches described in order to aid the reader in understanding their strengths and weaknesses. We hope this will make clear why we offer such recommendations in our discussion and conclusion. As we have up to this point, we will also make an effort to both present the mathematical formulations of the models described, as well as to provide an explanation of them in plain English.

@@

## Log ratio of coefficient of variation ($\textrm{ln}CVR$)
The coefficient of variation is the ratio of the standard deviation to the mean; therefore, comparison of the coefficient of variation between groups will identify whether standard deviations differ more, or less, than would be predicted by their difference in means where a mean-variance relationship is present. In essence, the coefficient of variation is a means of standardising the standard deviation against the mean such that the relative variation in an effect is expressed. The natural logarithm of the ratio between the coefficients of variation from two groups, the $\textrm{ln}CVR$ is thus a more generally applicable effect size statistic for examining variability between groups. Considering equations \@ref(eq:lnRR-eq) and \@ref(eq:lnVR-eq), the $\textrm{ln}CVR$ is given by:

@@

# Examples using resistance training studies
As noted, to facilitate understanding for those new to examination of variation, we provide primary examples of the approaches presented using data from RT studies included in a recent meta-analysis published in the *Journal of Sport Sciences* [@politoModeratorsStrengthGains2021]. Here we have used their list of included studies and re-extracted data from 111 of these[^12]. All analysis examples were performed in R (version 4.2.1, "Funny-Looking Kid", The R Foundation for Statistical Computing, 2022) using the **metafor** package [@viechtbauerConductingMetaAnalysesMetafor2010]. The extracted dataset, analysis scripts, models, data summaries, and supplementary materials are available on the Open Science Framework (https://osf.io/2h9ma/) or the GitHub repository (https://github.com/jamessteeleii/Meta-Analysis-of-Variation-in-Resistance-Training.git).

@@

[^13]: Regression analyses are likely familiar to most readers where in the simplest form they try to predict the value of some dependent variable from some independent variable(s). This can be extended to meta-analytic synthesis where the independent variables reflect characteristics associated with the effects included. For example, they may reflect characteristics of the sample in the study for which the effect was extracted such as age or sex, or they might reflect characteristics of the intervention received such as the dose or frequency of exposure.

@@

## Detecting the presence of interindividual response variation to resistance training intervention with absolute variance statistics
First we conducted a traditional SMD and $\textrm{ln}RR$ based effect size[^14] meta-analysis to explore the effects of RT interventions compared to controls for strength and hypertrophy (i.e., muscle mass/size) outcomes[^15]. Polito et al. [-@politoModeratorsStrengthGains2021] originally used a normal random-effects meta-analysis as described in the introduction. However, the data we extracted were hierarchical in nature. Thus, as opposed to there being only the assumption that studies are a random effect, due to there being multiple outcomes measured within each arm in the studies (i.e., intervention group(s) and control group, within each study), and that in some studies there were multiple interventions examined, there is the additional assumption that must be included that both the intervention groups and effects also come from overarching distributions. Thus a multilevel mixed-effects meta-analysis model [@vandennoortgateThreelevelMetaanalysisDependent2013] with cluster-robust variance estimation [@hedgesRobustVarianceEstimation2010] was used with random intercepts for study, arm[^16], and effect. This model then includes additional $\tau^2$ terms for each of the levels and assigns weights appropriately given this and the clustering of effects within arms within studies. We can therefore describe the overall model as:

@@

where $\hat\theta_{ijk}$ is the $k\textrm{th}$ effect size, here the SMD or $\textrm{ln}RR$, from the $j\textrm{th}$ arm ($j = 1,2,\cdots,N_{j}$; where $N_{j}$ is the number of arms) in the $i\textrm{th}$ study ($i = 1,2,\cdots,N_{i}$; where $N_{i}$ is the number of studies), $\theta$ is the intercept or overall mean of the effects, $\tau_{i}$ is the deviation from $\theta$ for the $i\textrm{th}$ study,  $\tau_{j}$ is the deviation for the $j\textrm{th}$ arm, $\tau_{k}$ is the deviation for the $k\textrm{th}$ effect, and $m_{ijk}$ is the sampling error for each effect size normally distributed with $\sigma^2_{\theta_{ijk}}$. This model is referred to as 'mixed' effects because of the presence of both fixed ($\theta$), and random ($\tau_{(1)i},\tau_{(2)j},\tau_{(3)k}$) effects[^17]. The main term in the model we are interested in is $\theta$ which is our estimate for the overall weighted average effect (i.e., $\hat\theta$).

[^17]: Technically then the *random* effects model presented earlier is also a *mixed* effects model. It is traditionally referred to as the random-effects model though.

@@

The pattern of results from our models examining SMDs (figure \@ref(fig:forest-SMD-plot)) were similar to those reported by Polito et al. [-@politoModeratorsStrengthGains2021], albeit with slightly lower estimates for both outcome types; possibly due to our use of a multilevel mixed-effects meta-analysis model that allowed for each individual effect size to be more appropriately weighted (the relative amount of heterogeneity between effects for each level is presented as the $I^2$ statistic). As might be expected, in comparison to non-training controls the RT interventions produced increases in both strength (SMD = `r SMD_RR_SDir_logVR$Estimate[1]` [95%CI: `r SMD_RR_SDir_logVR$Lower[1]` to `r SMD_RR_SDir_logVR$Upper[1]`]; $I^2_{study}$ = `r SMD_RR_SDir_logVR$I.2.study[1]`%, $I^2_{arm}$ = `r SMD_RR_SDir_logVR$I.2.arm[1]`%, $I^2_{effect}$ = `r SMD_RR_SDir_logVR$I.2.effect[1]`%) and hypertrophy outcomes (SMD = `r SMD_RR_SDir_logVR$Estimate[5]` [95%CI: `r SMD_RR_SDir_logVR$Lower[5]` to `r SMD_RR_SDir_logVR$Upper[5]`]; $I^2_{study}$ = `r SMD_RR_SDir_logVR$I.2.study[5]`%, $I^2_{arm}$ = `r SMD_RR_SDir_logVR$I.2.arm[5]`%, $I^2_{effect}$ = `r SMD_RR_SDir_logVR$I.2.effect[5]`%). Confidence intervals on the overall effects were precise for both outcomes though prediction intervals, indicating the range over which we might expect future estimates of effects to fall based on this evidence, for SMD estimates (see figure \@ref(fig:forest-SMD-plot)) were fairly wide and relative heterogeneity was fairly high mostly as a result of between-study variance (i.e., effects were more similar *within* studies and arms than *between* them).

@@

## Mean-variance relationships in muscular strength and hypertrophy
With meta-analytic models of variation we are not limited to solely exploring variation in responses to interventions (e.g., @halperinAccuracyPredictingRepetitions2022; @nuzzoMaximalNumberRepetitions2023). We can explore the relationships between variance in a number of outcomes and the impact of certain predictors on this in the form of meta-regression. For example, as noted, one possible predictor of variance is the mean itself. As such, we can model variance of each effect as the response itself with the mean of the effect as the predictor. The standard deviation is, however, bounded at zero and so in many cases it may not conform to assumptions of normality which are required for regression models (i.e., that the residuals, the difference between the estimated and actual data, are normally distributed). Therefore, we instead can use $\textrm{ln}\hat\sigma$, which is unbounded. In the following example we explore the mean-variance relationship in the pre-intervention scores for outcomes in the data set from Polito et al. [-@politoModeratorsStrengthGains2021].

>>>>

---
title: "Meta-Analysis of Variation in Sport and Exercise Science"
subtitle: "Examples of Application Within Resistance Training Research"
author: 
  - James Steele$^a$
  - James Fisher$^a$
  - Dave Smith$^b$
  - Brad Schoenfeld$^c$
  - Yefeng Yang$^d$
  - Shinichi Nakagawa$^d$
date: "14th September 2023"
abstract: "Meta-analysis has become commonplace within sport and exercise science for synthesising and summarising empirical studies. However, most research in the field focuses upon mean effects; particularly the effects of interventions to improve outcomes such as fitness or performance. It is thought that individual responses to interventions vary considerably. Hence, interest has increased in exploring *precision* or *personalised* exercise approaches. Not only is the mean often affected by interventions, but variances may also be impacted. Exploration of variances in studies such as randomised controlled trials (RCTs) can yield insight into interindividual heterogeneity in response to interventions and help determine generalisability of effects. Yet, larger samples sizes than those used for typical mean effects are required when probing variances. Thus, in a field with small samples such as sport and exercise science, exploration of variance through a meta-analytic framework is appealing. Despite the value of embracing and exploring variation alongside mean effects in sport and exercise science it is rarely applied to research synthesis through meta-analysis.We introduce and evaluate different effect size calculations along with models for meta-analysis of variation using relatable examples from resistance training RCTs."
thanks: |
  Address for correspondence: james.steele@solent.ac.uk
  Affiliations:
  $^a$Faculty of Sport, Health, and Social Sciences, Solent University, UK;
  $^b$Research Centre for Musculoskeletal and Sports Medicine, Manchester Metropolitan University, Manchester, UK;
  $^c$Health Sciences Department, CUNY Lehman College, Bronx, New York, USA;
  $^d$Evolution & Ecology Research Centre and School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, New South Wales, Australia
output: 
  bookdown::pdf_document2:
    extra_dependencies: ["flafter"]
bibliography: references.bib
csl: apa.csl
toc: false
header-includes:
   - \usepackage{caption}
   - \captionsetup[figure]{font=small}
   - \usepackage{lineno}
   - \linenumbers
---

@@

```
# Introduction
Although the quantitative synthesis of results across studies has existed since the 17th century [@plackettStudiesHistoryProbability1958], the modern-day term "meta-analysis" was coined by Gene Glass [-@glassPrimarySecondaryMetaAnalysis1976]. Since that time, the use of meta-analysis as a tool for the synthesis of research in sport and exercise science has increased considerably [@haggerMetaanalysis2022], with resistance training (RT) accounting for a considerable proportion of this growth (figure \@ref(fig:trend-plot)). Accordingly, throughout the paper we use RT studies as a hopefully familiar example for sport and exercise science researchers. \textcolor{red}{However, to begin we provide a conceptual overview of meta-analyses and effect sizes as typically used within sport and exercise science.}

@@

\textcolor{red}{There are two popular models}[^1] \textcolor{red}{used for meta-analysis: the fixed-effect model and the random-effects model} [@borensteinBasicIntroductionFixedeffect2010]. \textcolor{red}{The fixed effect model assumes that there is one true effect size}[^2] \textcolor{red}{that each study included in the meta-analysis has estimated and that any differences in the estimates between individual studies are due to only sampling error. This essentially means that there is a single common effect which is fixed across studies and each study takes samples of individuals from the population to estimate this effect. We can express this model in the following formula:}

[^1]: \textcolor{red}{To clarify language here for those unfamiliar, the term and concept *model* is used commonly in statistics. A statistical model essentially is specification of what we think the data generating process might be for a given situation. In the context of meta-analyses the data are usually the individual effects that we have extracted from studies i.e., the results of each study. The model, in mathematical formulae, is intended to approximate the processes that we assume led to the generation of the data.}

[^2]: \textcolor{red}{*Effect size* is an agnostic term used for a family of statistics which communicate the strength of a given 'effect' resulting from research. This includes descriptive statistics ranging from mean raw values to correlation coeffients and everything in between} [@caldwellCaseDefaultEffect2020] \textcolor{red}{including, as we shall see, statistics describing variation.}

@@

\textcolor{red}{where $\hat\theta_{i}$ is the $i\textrm{th}$ effect size ($i = 1,2,\cdots,N_{i}$; where $N_{i}$ is the number of studies and thus effect sizes), $\theta$ is the intercept or overall mean (i.e., the fixed-effect), and $m_{i}$ is the sampling error for each effect size normally distributed with $\sigma^2_{m_{i}}$. Studies with smaller standard errors of their effect estimates have smaller sampling errors and so these higher precision estimates are given greater weight in the model. The weighting given to each study is calculated as:}

@@

\textcolor{red}{Where $s_{i}$ is the standard deviation for the effect estimate and thus $s^2_{i}$ is the variance. It is referred to as inverse-variance weighting. Then, the overall weighted mean effect estimate from the model is then calculated as:}

@@

\textcolor{red}{Contrastingly, the random effects model does not make the assumption that there is only one fixed-effect. Instead it allows for the true effects that each study estimates to differ. Each study may share a common underlying effect size but, due to differences between studies in factors such as population characteristics, the manner in which outcomes are operationalised, or subtle differences in intervention and context to name a few, it is possible that the actual effect being estimated by each study differs. The assumption of the random-effects model then is that the studies included estimate effects that come from a larger population of effects determined by the inclusion criteria for studies included. So, the model assumes that the studies included reflect a random sample of all possible permutations of study and the effects they estimate from this population distribution of studies and effects. Hence, in the fixed-effect model there is one effect and it is assumed to be *fixed* across studies, whereas in the random-effects model there are many and we examine an assumed *random* sample of them. We can express this model in the following formula:}

@@

\textcolor{red}{where $\hat\theta_{i}$ is the $i\textrm{th}$ effect size ($i = 1,2,\cdots,N_{i}$; where $N_{i}$ is the number of studies and thus effect sizes), $\theta_i$ is the intercept or overall mean of the effects across each study, $\tau_{i}$ is the deviation from $\theta_{i}$ for the $i\textrm{th}$ study (i.e., the random-effects), and $m_{i}$ is the sampling error for each effect size normally distributed with $\sigma^2_{m_{i}}$. Essentially this model assumes that each individual study estimates an effect and there will be some sampling error in estimating it. But, the effects each study estimates comes from an overall distribution of true effects with a mean value (i.e., $\theta_{i}$). The weighting in any model employing random effects such as this requires a different approach as the variance of the distribution of the effect sizes the model assumes, known as $\tau^2$, must also be estimated}[^3]. \textcolor{red}{This essentially describes the heterogeneity between the effects included. Once this has been estimated we can calculate a weight that is adjusted for the random effects for each effect as:}

[^3]: \textcolor{red}{This estimation can be done using a variety of methods and is an area of ongoing investigation as to how different methods perform. This is beyond the scope of this paper to discuss. We note however that the models we present all utilise Restricted Maximum Likelihood estimation.}

@@

\textcolor{red}{Then the average of the distribution of effects can be calculated as per equation} \@ref(eq:fixed-average-eq) \textcolor{red}{substituting $w_{i}$ for $w^*_{i}$.}

\textcolor{red}{Although historically fixed effect models were commonplace in the field of sport and exercise science} [@haggerMetaanalysisSportExercise2006] \textcolor{red}{nowadays the random effects model is more often employed} [@haggerMetaanalysis2022]. \textcolor{red}{This is likely due to the fact that direct (i.e., exact) replication of studies is rare}[^4] \textcolor{red}{and instead studies often explore similar effects across varying moderating factors such as those noted above.}

[^4]: \textcolor{red}{Hence current efforts to conduct direct replications (see https://ssreplicationcentre.com/).}

As with many other fields [@nakagawaMetaanalysisVariationEcological2015; @usuiMetaanalysisVariationSuggests2021; @millsDetectingHeterogeneityIntervention2021] \textcolor{red}{likely the most common aim in meta-analysis in sport and exercise science, and indeed primary empirical research too, is to estimate the *effect* of an independent variable upon some dependent outcome variable. The dependent variable is often the mean of a measurement and the independent variable is often a categorical grouping}; for example, the comparison of an intervention group(s) and a control group, the comparison of intervention groups between one another, or comparison between non-manipulated categories such as biological sex. Indeed, a recent umbrella review [@bernardez-vazquezResistanceTrainingVariables2022] of meta-analyses in RT identified 14 studies examining the manipulation of RT intervention variables (i.e., the comparison of one intervention to another whereby a variable in the intervention was manipulated) on hypertrophy outcomes, all of which focused on the comparison of mean changes between different intervention groups.

\textcolor{red}{Often, due to the varying operationalisations used for broad outcome concepts}[^5]\textcolor{red}{, an effect size is used which is "standardised" across studies to allow for their synthesis.} Most commonly, a magnitude based[^6] effect size statistic [@caldwellCaseDefaultEffect2020], the standardised mean difference (SMD), is used to compare means between groups or conditions. This statistic is usually a version of Cohen's *d* [@cohenStatisticalPowerAnalysis1988], or its bias-corrected[^7] metric referred to as Hedges' *g* [@hedgesStatisticalMethodsMetaAnalysis2014; @borensteinIntroductionMetaAnalysis2021; @nakagawaEffectSizeConfidence2007][^8]. The SMD, and its sampling variance, $s^2_{SMD}$ are given by:

[^5]: \textcolor{red}{For example, *strength* might be examined in different studies using different operationalisations including one repetition maximum testing or maximum voluntary contractions. Or the same operationalisations may be employed but different exercises such as the squat or bench press.}

@@

[^7]: \textcolor{red}{For those unfamiliar with the terminology, an estimator for a statistic is unbiased if it produces parameter estimates that are on average correct. Thus a bias corrected statistic is one which would be biased without the correction applied, but otherwise has been shown to be unbiased.}

[^8]: We will refer to both merely as the SMD throughout the manuscript for simplicity and note that throughout when reporting a 'SMD' we are reporting the bias-corrected version. \textcolor{red}{We also note that another magnitude based effect size, Glass' $\Delta$, is commonly recommended as it is the simplest form of SMD though makes assumptions about the impact of the intervention having no effect on the denominator (i.e., variance;} @caldwellCaseDefaultEffect2020).

@@

Comparison of heterogeneity in responses, such as post-scores or change scores to interventions, are not the only possible use of statistical methods for comparing variances. For example, in other fields such as ecology there have been calls to shift focus of analysis onto the exploration of dispersion of traits between groups in non-experimental or intervention designs [@nakagawaMeanStrikesBack2012]. Some recent examples from sport and exercise science, and RT in particular, include primary research exploring between-participant acute response variation for the purposes of identifying methods[^9] to reduce RT stimulus heterogeneity, [@exnerDoesPerformingResistance2022] as well as a meta-analysis exploring between-participant heterogeneity of accuracy in predicting proximity to task failure during RT [@halperinAccuracyPredictingRepetitions2022] \textcolor{red}{and in the number of repetitions that can be performed at different percentages of one repetition maximum} [@nuzzoMaximalNumberRepetitions2023].

@@

\textcolor{red}{Although ultimately we will recommend the thoughtful consideration of assumptions for certain approaches regarding what effect size statistics and models to employ in examining variation through use of meta-analysis, we provide examples throughout using all approaches described in order to aid the reader in understanding their strengths and weaknesses. We hope this will make clear why we offer such recommendations in our discussion and conclusion. As we have up to this point, we will also make an effort to both present the mathematical formulations of the models described, as well as to provide an explanation of them in plain English.}

@@

## Log ratio of coefficient of variation ($\textrm{ln}CVR$)
The coefficient of variation is the ratio of the standard deviation to the mean; therefore, comparison of the coefficient of variation between groups will identify whether standard deviations differ more, or less, than would be predicted by their difference in means where a mean-variance relationship is present. \textcolor{red}{In essence, the coefficient of variation is a means of standardising the standard deviation against the mean such that the relative variation in an effect is expressed.} The natural logarithm of the ratio between the coefficients of variation from two groups, the $\textrm{ln}CVR$ is thus a more generally applicable effect size statistic for examining variability between groups. Considering equations \@ref(eq:lnRR-eq) and \@ref(eq:lnVR-eq), the $\textrm{ln}CVR$ is given by:

@@

# Examples using resistance training studies
As noted, to facilitate understanding for those new to examination of variation, we provide primary examples of the approaches presented using data from RT studies included in a recent meta-analysis published in the *Journal of Sport Sciences* [@politoModeratorsStrengthGains2021]. Here we have used their list of included studies and re-extracted data from 111 of these[^12]. All analysis examples were performed in R (version 4.2.1, "Funny-Looking Kid", The R Foundation for Statistical Computing, 2022) using the **metafor** package [@viechtbauerConductingMetaAnalysesMetafor2010]. The extracted dataset, analysis scripts, models, data summaries, and supplementary materials are available on the Open Science Framework (https://osf.io/2h9ma/) \textcolor{red}{or the GitHub repository (https://github.com/jamessteeleii/Meta-Analysis-of-Variation-in-Resistance-Training.git)}.

@@

[^13]: \textcolor{red}{Regression analyses are likely familiar to most readers where in the simplest form they try to predict the value of some dependent variable from some independent variable(s). This can be extended to meta-analytic synthesis where the independent variables reflect characteristics associated with the effects included. For example, they may reflect characteristics of the sample in the study for which the effect was extracted such as age or sex, or they might reflect characteristics of the intervention received such as the dose or frequency of exposure.}

@@

## Detecting the presence of interindividual response variation to resistance training intervention \textcolor{red}{with absolute variance statistics}
First we conducted a traditional SMD and $\textrm{ln}RR$ based effect size[^14] meta-analysis to explore the effects of RT interventions compared to controls for strength and hypertrophy (i.e., muscle mass/size) outcomes[^15]. Polito et al. [-@politoModeratorsStrengthGains2021] originally used a normal random-effects meta-analysis as described in the introduction. However, the data we extracted were hierarchical in nature. \textcolor{red}{Thus, as opposed to there being only the assumption that studies are a random effect, due to there being multiple outcomes measured within each arm in the studies (i.e., intervention group(s) and control group, within each study), and that in some studies there were multiple interventions examined, there is the additional assumption that must be included that both the intervention groups and effects also come from overarching distributions.} Thus a multilevel mixed-effects meta-analysis model [@vandennoortgateThreelevelMetaanalysisDependent2013] with cluster-robust variance estimation [@hedgesRobustVarianceEstimation2010] was used with random intercepts for study, arm[^16], and effect. \textcolor{red}{This model then includes additional $\tau^2$ terms for each of the levels and assigns weights appropriately given this and the clustering of effects within arms within studies. We can therefore describe the overall model as:}

@@

\textcolor{red}{where $\hat\theta_{ijk}$ is the $k\textrm{th}$ effect size, here the SMD or $\textrm{ln}RR$, from the $j\textrm{th}$ arm ($j = 1,2,\cdots,N_{j}$; where $N_{j}$ is the number of arms) in the $i\textrm{th}$ study ($i = 1,2,\cdots,N_{i}$; where $N_{i}$ is the number of studies), $\theta$ is the intercept or overall mean of the effects, $\tau_{i}$ is the deviation from $\theta$ for the $i\textrm{th}$ study,  $\tau_{j}$ is the deviation for the $j\textrm{th}$ arm, $\tau_{k}$ is the deviation for the $k\textrm{th}$ effect, and $m_{ijk}$ is the sampling error for each effect size normally distributed with $\sigma^2_{\theta_{ijk}}$. This model is referred to as 'mixed' effects because of the presence of both fixed ($\theta$), and random ($\tau_{(1)i},\tau_{(2)j},\tau_{(3)k}$) effects}[^17]. \textcolor{red}{The main term in the model we are interested in is $\theta$ which is our estimate for the overall weighted average effect (i.e., $\hat\theta$).}

[^17]: \textcolor{red}{Technically then the *random* effects model presented earlier is also a *mixed* effects model. It is traditionally referred to as the random-effects model though.}

@@

The pattern of results from our models examining SMDs (figure \@ref(fig:forest-SMD-plot)) were similar to those reported by Polito et al. [-@politoModeratorsStrengthGains2021], albeit with slightly lower estimates for both outcome types; possibly due to our use of a multilevel mixed-effects meta-analysis model that allowed for each individual effect size to be more appropriately weighted (the relative amount of heterogeneity between effects for each level is presented as the $I^2$ statistic). As might be expected, in comparison to non-training controls the RT interventions produced increases in both strength (SMD = `r SMD_RR_SDir_logVR$Estimate[1]` [95%CI: `r SMD_RR_SDir_logVR$Lower[1]` to `r SMD_RR_SDir_logVR$Upper[1]`]; $I^2_{study}$ = `r SMD_RR_SDir_logVR$I.2.study[1]`%, $I^2_{arm}$ = `r SMD_RR_SDir_logVR$I.2.arm[1]`%, $I^2_{effect}$ = `r SMD_RR_SDir_logVR$I.2.effect[1]`%) and hypertrophy outcomes (SMD = `r SMD_RR_SDir_logVR$Estimate[5]` [95%CI: `r SMD_RR_SDir_logVR$Lower[5]` to `r SMD_RR_SDir_logVR$Upper[5]`]; $I^2_{study}$ = `r SMD_RR_SDir_logVR$I.2.study[5]`%, $I^2_{arm}$ = `r SMD_RR_SDir_logVR$I.2.arm[5]`%, $I^2_{effect}$ = `r SMD_RR_SDir_logVR$I.2.effect[5]`%). Confidence intervals on the overall effects were precise for both outcomes though prediction intervals, \textcolor{red}{indicating the range over which we might expect future estimates of effects to fall based on this evidence, for SMD estimates (see figure }\@ref(fig:forest-SMD-plot)) \textcolor{red}{were fairly wide and relative heterogeneity was fairly high mostly as a result of between-study variance (i.e., effects were more similar *within* studies and arms than *between* them).}

@@

## Mean-variance relationships in muscular strength and hypertrophy
With meta-analytic models of variation we are not limited to solely exploring variation in responses to interventions (e.g., @halperinAccuracyPredictingRepetitions2022; @nuzzoMaximalNumberRepetitions2023). We can explore the relationships between variance in a number of outcomes and the impact of certain predictors on this in the form of meta-regression. For example, as noted, one possible predictor of variance is the mean itself. As such, we can model variance of each effect as the response itself with the mean of the effect as the predictor. The standard deviation is, however, bounded at zero and so in many cases it may not conform to assumptions of normality which are required for regression models \textcolor{red}{(i.e., that the residuals, the difference between the estimated and actual data, are normally distributed).} Therefore, we instead can use $\textrm{ln}\hat\sigma$, which is unbounded. In the following example we explore the mean-variance relationship in the pre-intervention scores for outcomes in the data set from Polito et al. [-@politoModeratorsStrengthGains2021].
